---
title: Intro to R - Exercises
author: "Student Name"
date: February 2, 2026
format: docx
---

The exercises in this document follow along closely with the [Introduction to R and RStudio](https://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro.html), [Seeking Help](https://swcarpentry.github.io/r-novice-gapminder/instructor/03-seeking-help.html) and [Data Structures](https://swcarpentry.github.io/r-novice-gapminder/instructor/04-data-structures-part1.html) modules of [R for Reproducible Scientific Analysis](https://swcarpentry.github.io/r-novice-gapminder/), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).

::: callout-tip
## Key Points

-   Intro to R
    -   R has the usual arithmetic operators and mathematical functions.
    -   Use `<-` to assign values to variables.
    -   Use `ls()` to list the variables in a program.
    -   Use `rm()` to delete objects in a script.
    -   Use `install.packages()` to install packages and `library()` to load them.
-   Seeking Help
    -   Use `help()` to get online help in R.
-   Data Structures
    -   Use read.csv to read tabular data in R.
    -   The basic data types in R are double, integer, complex, logical, and character.
    -   Data structures such as data frames or matrices are built on top of lists and vectors, with some added attributes.
:::

## Intro to R

#### Challenge 1

Which of the following are valid R variable names?

```         
min_height
max.height
_age
.mass
MaxLength
min-length
2widths
celsius2kelvin
```

<!-- Challenge 1 answer begin -->

-   Valid
-   Invalid

<!-- Challenge 1 answer end -->

#### Challenge 2

What will be the value of each variable after each statement in the following lines of code?

```{r}
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20
```

<!-- Challenge 2 answer begin -->

<!-- Challenge 2 answer end -->

#### Challenge 3

Run the code from the previous challenge, and write a command to compare `mass` to `age`. Is mass larger than age?

<!-- Challenge 3 answer begin -->

```{r}
# comparison code goes here
```

<!-- Challenge 3 answer end -->

#### Challenge 4

Clean up your working environment by deleting the mass and age variables.

<!-- Challenge 4 answer begin -->

```{r}
# code goes here
```

<!-- Challenge 4 answer end -->

#### Challenge 5

Install the following packages: `ggplot2`, `dplyr`, `gapminder`

<!-- Challenge 5 answer begin -->

```{r}
# this will fail if the required packages are not installed
library(ggplot2)
library(dplyr)
library(gapminder)
```

<!-- Challenge 5 answer end -->

## Seeking Help

#### Challenge 6

Look at the help page for `c`. What kind of vector do you expect will be created if you evaluate the following?

```{r}
c(1, 2, 3)
c('d', 'e', 'f')
c(1, 2, 'f')
```

<!-- Challenge 6 answer begin -->

Your answer goes here

<!-- Challenge 6 answer end -->

#### Challenge 7

Look at the help for the `paste` function. You will need to use it later. What's the difference between the `sep` and `collapse` arguments?

<!-- Challenge 7 answer begin -->

Your answer goes here

<!-- Challenge 7 answer end -->

#### Challenge 8

Use help to find a function (and its associated parameters) that you could use to load data from a tabular file in which columns are delimited with "\t\" (tab) and the decimal point is a"." (period). This check for decimal separator is important, especially if you are working with international colleagues, because different countries have different conventions for the decimal point (i.e. comma vs period).

::: {.callout-note collapse="true"}
# Hint

Use `??"read table"` to look up functions related to reading in tabular data.
:::

<!-- Challenge 8 answer begin -->

Your answer goes here

<!-- Challenge 8 answer end -->

## Data Structures

#### Challenge 9

An important part of every data analysis is cleaning the input data. If you know that the input data is all of the same format, (e.g. numbers), your analysis is much easier! Clean the cat data set from the chapter about type coercion.

Use the starter code provided here and complete the tasks below.

<!-- Challenge 9 answer begin -->

```{r}
# Read data (make sure you are in the correct working directory here!)
cats <- data.frame(coat = c("calico", "black", "tabby", "tabby"),
           weight = c(2.1, 5.0, 3.2, "2.3 or 2.4"),
           likes_catnip = c(1, 0, 1, 1))

# 1. Print the data
## your command goes here ##

# 2. Show an overview of the table with all data types
## your command goes here ##

# 3. The "weight" column has the incorrect data type ## your comment goes here ##.
#    The correct data type is: ## your comment goes here ##

# 4. Correct the 4th weight data point with the mean of the two given values
cats$weight[4] <- 2.35

#    print the data again to see the effect
cats

# 5. Convert the weight to the right data type
cats$weight <- ## your command goes here ##

#    Calculate the mean to test yourself
mean(cats$weight)

# If you see the correct mean value (and not NA), you did the exercise
# correctly!
```

<!-- Challenge 9 answer end -->

#### Challenge 10

Make a vector with the numbers 1 through 26 (save this vector to the variable `x`). Then, multiply the vector by 2.

<!-- Challenge 10 answer begin -->

```{r}
# your commands go here
```

<!-- Challenge 10 answer end -->

#### Challenge 11

```{r}
named_vector <- c(a = 1, b = 2, c = 3)
```

What is the data type of the names of `named_vector`? What function(s) can we use to check this?

<!-- Challenge 11 answer begin -->

```{r}
# your command goes here
```

<!-- Challenge 11 answer end -->

#### Challenge 12

Instead of just changing some of the names a vector/list already has, you can also set all names of an object by writing code like (replace ALL CAPS text):

```         
names( OBJECT ) <-  CHARACTER_VECTOR
```

Create a vector that gives the number for each letter in the alphabet! R has a built-in object called `LETTERS`. It is a 26-character vector, from A to Z.

<!-- Challenge 12 answer begin -->

```{r}
# Generate a vector called letter_no with the sequence of numbers from 1 to 26!
letter_no <- ## your command goes here ##

# Set the names of the number sequence to the 26 letters in `LETTERS`.
## your command goes here ##

# Test yourself by calling `letter_no["B"]`, which should give you the number 2!
letter_no["B"]
```

<!-- Challenge 12 answer end -->

#### Challenge 13

There are several subtly different ways to call variables, observations and elements from `data.frames`. Try out these examples and explain what is returned by each one. *Hint*: Use the function `typeof()` to examine what is returned in each case.

<!-- Challenge 13 answer begin -->

```{r}
cats[1]
```

Explain output here.

```{r}
cats[[1]]
```

Explain output here.

```{r}
cats$coat
```

Explain output here.

```{r}
cats["coat"]
```

Explain output here.

```{r}
cats[1, 1]
```

Explain output here.

```{r}
cats[, 1]
```

Explain output here.

```{r}
cats[1, ]
```

<!-- Challenge 13 answer end -->

#### Challenge 14

Given `matrix_example`,

```{r}
matrix_example <- matrix(0, ncol=6, nrow=3)
```

what do you think will be the result of `length(matrix_example)`? Try it. Were you right? Why / why not?

<!-- Challenge 14 answer begin -->

Explain here.

<!-- Challenge 14 answer end -->

#### Challenge 15

Make another matrix, this time containing the numbers 1:50, with 5 columns and 10 rows. Did the matrix function fill your matrix by column, or by row, as its default behavior? See if you can figure out how to change this. (hint: read the documentation for matrix!)

<!-- Challenge 15 answer begin -->

```{r}
# using matrix with default `matrix` parameters
## your command goes here

# modify the previous command to fill in the opposite order as above
## your command goes here
```

<!-- Challenge 15 answer end -->

#### Challenge 16

Create a list of length two containing a character vector for each of the sections in this part of the workshop:

-   Data types
-   Data structures

Populate each character vector with the names of the data types and data structures we've seen so far.

<!-- Challenge 16 answer begin -->

```{r}
## you command goes here
```

<!-- Challenge 16 answer end -->

#### Challenge 17

Consider the R output of the matrix below:

```         
     [,1] [,2]
[1,]    4    1
[2,]    9    5
[3,]   10    7
```

What was the command that generated this output? Examine each command and try to figure out the correct one before typing them. Think about what matrices the other commands will produce.

1.  `matrix(c(4, 1, 9, 5, 10, 7), nrow = 3)`
2.  `matrix(c(4, 9, 10, 1, 5, 7), ncol = 2, byrow = TRUE)`
3.  `matrix(c(4, 9, 10, 1, 5, 7), nrow = 2)`
4.  `matrix(c(4, 1, 9, 5, 10, 7), ncol = 2, byrow = TRUE)`

<!-- Challenge 17 answer begin -->

```{r}
# your codmmand goes here
```

<!-- Challenge 17 answer end -->