---
title: ggplot2 - Exercises
author: "Student Name"
date: February 23, 2026
format: docx
---

These exercises follow along closely with the [ggplot2](https://swcarpentry.github.io/r-novice-gapminder/08-plot-ggplot2.html) module of [R for Reproducible Research](https://swcarpentry.github.io/r-novice-gapminder/index.html), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).

## ggplot2

### Challenge 1

Modify the following code so that the figure shows how life expectancy has changed over time.

```{r}
library(ggplot2)
library(gapminder)

ggplot(data = gapminder, mapping = aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

### Challenge 2

In the previous examples and challenge we've used the aes function to tell the scatterplot geom about the x and y locations of each point. Another aesthetic property we can modify is the point color. Modify the code from the previous challenge to color the points by the `continent` column.

```{r}
# your code goes here
```

What trends do you see in the data? Are they what you expected?

*your answer goes here*

### Challenge 3

Switch the order of the point and line layers from the previous example.

```{r}
# your code goes here
```

What happened?

*your answer goes here*

### Challenge 4

Modify the color and size of the points on the point layer in the figure below.

-   Hint: do not use the aes function.
-   Hint: the equivalent of linewidth for points is size

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = "lm")
```

Now modify the points to be a different shape, and color them by continent as well.

```{r}
ggplot(gapminder, aes(gdpPercap, lifeExp)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() +
  geom_smooth(method = "lm")
```

### Challenge 5

Generate boxplots to compare life expectancy between the different continents during the available years (i.e. there should be a separate panel for each year in the data set).

Bonus:

-   Add jittered points to the boxplot to show the distribution of the data
-   Add violin plots to the boxplot to show the distribution of the data
-   Rename the y-axis and remove x-axis labels

```{r}
# your code goes here
```

### Challenge 6

Modify the following plot such that it is easier to read and interpret. You can modify the colors, point size, point shape, line type, line width, and/or the theme. You may also want to refer to the [ColorBrewer](https://colorbrewer2.org/) palettes to choose a color scheme that is colorblind friendly and works well in black and white.

```{r}
library(RColorBrewer)

ggplot(gapminder, aes(year, lifeExp, group=country, color=continent)) +
  geom_jitter(height = 0, alpha = 0.5) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")
```
