---
title: Vectorization and Functions - Exercises
author: "Student Name"
date: March 2, 2026
format: docx
---

These exercises follow along closely with the [Vectorization](https://swcarpentry.github.io/r-novice-gapminder/09-vectorization.html) and [Functions](https://swcarpentry.github.io/r-novice-gapminder/10-functions.html) modules of [R for Reproducible Research](https://swcarpentry.github.io/r-novice-gapminder/index.html), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).


## Vectorization

### Challenge 1

Practice working with vectorized data on the `pop` column of the `gapminder` dataset.

Make a new column in the `gapminder` data frame that contains population in units of millions of people. Check the head or tail of the data frame to make sure it worked.

```{r}
# your code goes here
```

### Challenge 2

On a single graph, plot population, in millions, against year, for all countries. Do not worry about identifying which country is which.

```{r}
# your code goes here
```

Repeat the exercise, graphing only for China, India, and Indonesia. Again, do not worry about which is which.

```{r}
# your code goes here
```

### Challenge 3

Given the following matrix:

```{r}
(m <- matrix(1:12, nrow=3, ncol=4))
```

Write down what happens when you run the following expressions? Did you get the output you expected?

-   `m ^ -1`
-   `m * c(1, 0, -1)`
-   `m > c(0, 20)`
-   `m * c(1, 0, -1, 2)`

*your answers go here*

### Challenge 4

We're interested in looking at the sum of the following sequence of fractions:

```
x = 1/(1^2) + 1/(2^2) + 1/(3^2) + ... + 1/(n^2)
```

This would be tedious to type out, and impossible for high values of `n`. Use vectorisation to compute `x` when `n=100`.

Start by writing a `for` loop to compute this sum. Sometimes this can be helpful when trying to figure out how to vectorize a calculation.

```{r}
# for loop code goes here
```

Now, with this in mind, vectorize this `for` loop and calculate the result when `n=10,000`?

```{r}
# vectorized code goes here
```

## Functions

### Challenge 5

Write a function called `celsius_to_kelvin()` that takes a temperature in Kelvin and returns that temperature in Celsius.

Hint: To convert from Celsius to Kelvin, we add 273.15

```{r}
#' celsius_to_kelvin
#' Convert degrees Celsius to degrees Kelvin
#' 
#' @param temp Temperature in degrees Celsius
#' 
#' @return The temperature in degrees Kelvin
# your code goes here
```

### Challenge 6

Write and test a function called `fahr_to_kelvin` that combines `fahr_to_celsius` and `celsius_to_kelvin`.

```{r}
# From the lecture notes:
#' fahr_to_celsius
#' 
#' Convert degrees Fahrenheit to Celsius
#' 
#' @param temp Vector of temperatures in degrees Fahrenheit
#' 
#' @return Vector of temperatures in degrees Celsius
fahr_to_celsius <- function(temp) {
  (temp - 32) * (5 / 9)
}


#' fahr_to_kelvin
#' Convert degrees Fahrenheit to degrees Kelvin
#' 
#' @param temp A vector of temperatures in degrees Fahrenheit
#' 
#' @return A vector of temperatures in degrees Kelvin
# your code goes here
```

### Challenge 7

Use defensive programming to ensure that our functions throw a errors immediately if the argument `temp` is specified inappropriately. Do we *need* to check `temp` for `fahr_to_kelvin`? Why? What should we consider when deciding whether to check a particular input?

```{r}
# your code goes here
```

### Challenge 8

Storing our functions in a separate file is helpful for a lot of reasons. Copy all temperature conversion functions to a separate file and load it here with `source`.

```{r}
# your code goes here

# functions still work
fahr_to_celsius(212)
fahr_to_kelvin(32)
```

### Challenge 9

Test out `calcGDP` by calculating the GDP for New Zealand in 1952 and 1987. How do these values differ?

```{r}
# your code goes here
```

### Challenge 10

The `paste()` function can be used to combine text together, e.g:

```{r}
best_practice <- c("Write", "programs", "for", "people", "not", "computers")
paste(best_practice, collapse=" ")
```

Write a function called `fence()` that takes two vectors as arguments, called `text` and `wrapper`, and prints out the text wrapped with the wrapper. So, the output of `fence("Hello!")` would be:

```
***Hello!***
```

:::{.callout-note}
#### Note

The `paste()` function has an argument called `sep`, which specifies the separator between text. The default is a space: `" "`. The default for `paste0()` is no space, `""`.

:::

```{r}
#' fence
#' Print out text wrapped inside a wrapper
#' 
#' @param text A character vector
#' @param wrapper A character value (default is "***")
#' 
#' @return A character vector wrapped in a the wrapper text
# your code goes here
```

### Challenge 11

For our final challenge, we will revisit the p-series we introduced in Challenge 4 above. Create a new function, `p_series`, that will caclulate the sum of the following series:

```
x = 1/(1^p) + 1/(2^p) + ... + 1/(n^p)
```

```{r}
#' p_series
#' Calculate the sum of a p-series
#' 
#' @param n Integer value defining how many terms to sum up
#' @param p Integer value defining the power term
#' 
#' @return The following sum is returned: 1/(1^p) + 1/(2^p) + ... + 1/(n^p)

# your code goes here
```