---
title: Data Frames and the Tidyverse - Exercises
author: "Student Name"
date: March 16, 2026
format: docx
---

These exercises follow along closely with the [Data Frame Manipulation with dplyr](https://swcarpentry.github.io/r-novice-gapminder/12-dplyr.html) and [Data Frame Manipulation with tidyr](https://swcarpentry.github.io/r-novice-gapminder/13-tidyr.html) modules of [R for Reproducible Research](https://swcarpentry.github.io/r-novice-gapminder/index.html), provided by the [Software Carpentry](https://software-carpentry.org/) community under the creative commons license, [CC-BY 4.0](https://creativecommons.org/licenses/by/4.0/). For more information see [The Carpentries](https://carpentries.org/).

```{r}
#| include: false
# We typically will want to load our packages and perhaps perform some preliminary data wrangling
# at the top of our quarto document.
# Adding `#| include: false` will ignore any messages printed to the screen
# when these packages are loaded, resulting in a cleaner document.
```

## `dplyr`

### Challenge 1

Write a single command (which can span multiple lines and includes pipes) that will produce a data frame that has the African values for `lifeExp`, `country` and `year`, but not for other continents.

```{r}
# your code goes here
```

How many rows does your data frame have and why?

*your answer goes here*

### Challenge 2

Calculate the average life expectancy per country.

```{r}
# your answer goes here
```

Which has the longest average life expectancy and which has the shortest average life expectancy?

*Your answer goes here*

### Challenge 3

Modify the following code to use `mutate`.

```{webr-r}
library(ggplot2)

# Filter countries located in the Americas
americas <- gapminder[gapminder$continent == "Americas", ]
# Make the plot
ggplot(data = americas, mapping = aes(x = year, y = lifeExp)) +
  geom_line() +
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle = 45))
```


## `tidyr`

### Challenge 4

Is gapminder a purely long, purely wide, or some intermediate format? Why?

*your answer goes here*

### Challenge 5

Using the longer version of gapminder, calculate the mean life expectancy, population, and gdpPercap for each continent. Hint: use the `group_by` and `summarize` functions we learned in the `dplyr` lesson. Can you do it using pipes?

```{r}
read.csv('https://swcarpentry.github.io/r-novice-gapminder/data/gapminder_wide.csv') |>

  # this makes our data a little too long
  pivot_longer(cols = c(starts_with('pop'), starts_with('lifeExp'), starts_with('gdpPercap')),
               names_to = "obstype_year",
               values_to = "obs_values") |>
  
  separate(obstype_year, into = c('obs_type', 'year'), sep = "_") |>
  mutate(year = as.integer(year))

  # your code goes here
```

### Challenge 6

Take the `gap_wide` data frame 1 step further and create a `gap_ludicrously_wide` data frame by pivoting over countries, year and the 3 metrics? Hint this new data frame should only have 5 rows.

```{r}
# your code goes here
```
